Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-11 paper2.qmd
section:5; process:write,formulas,code

<!-- commands for d-separation -->
\newcommand{\dsep}{\:\bot\:}
\newcommand{\ndsep}{\:\not\bot\:}
\newcommand{\cond}{\:|\:}

@@

<!-- 1. The SCMs -->
Before incorporating the first theoretical principles of Thurstone's theory, it is essential to further define SCMs. SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other endogenous and exogenous variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

@@

@@

@@

@@

<!-- 1. start by defining the comparison vector  -->
As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model (see equation ([-@eq-mat27])). The vector contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $S$ in equation ([-@eq-mat26]).
$$
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat27}

<!-- 2. What is the effect of the comparison mechanism -->
The [DAG -@fig-cj14_dag] also integrates the comparison mechanism $C$ into the conceptual-population model, which ultimately derives the final comparative outcome $O_{R}$. First, it shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism. Second, it depicts the *comparison mechanism* $C$ as a causal factor influencing the final comparative outcome $O_{R}$. As with the sampling mechanism, a square encloses $C$ to indicate that it is a conditioned variable--that is, one that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*. In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli. Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp. 267].

@@

Notably, [DAG -@fig-cj14_dag] shows that $C$ is independent of all other variables in the model. This independence implies that the conceptual model represented by the DAG applies exclusively to Random Allocation Comparative Designs [@Bramley_2015] or Incomplete Block Designs [@Lawson_2015]. In these designs, every repeated judgment has an equal probability of being included in the sample.

@@

<!-- 3. preparing to transform the conceptual model into an statistical one -->
Finally, without loss of generality, researchers can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15. This reformulation rests on two principles. First, because the sampling and comparison mechanism vectors $S$ and $C$ are binary, researchers can integrate them into the model using element-wise products. This approach eliminates the need to define the structural equations $f_{S}$ and $f_{C}$. Second, since only the final comparative outcome $O_{R}$ is observed, it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$, defined by the structural equation $f_{O}$, also applies to this outcome. This assumption removes the necessity of describing the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$.

<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population. For example, if I have the distribution of the population outcome, and it is a bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->

<!-- 4. segway for next section -->
In summary, extending Thurstone's general form allows researchers to address issues with the BTL model, such as accounting for judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2). However, it does not address concerns about stimuli' equal dispersions discussed in @sec-theory-issue1a. Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we develop in the next section.

# Abandoning the BTL model {#sec-statistical}

<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
<!-- recording: Sven and Tine 25.03.28; time: 00:33:55 - 00:37:50 -->
<!-- recording: Sven and Tine 25.03.31; time: 00:38:50 - 00:47:00 -->

<!-- Deriving a statistical model to address violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a) becomes straightforward when using SCMs. Because the structural equations $F$ in the [SCM -@fig-cj15_scm] encode probabilistic information, the SCM can represent the joint distribution of the variables more efficiently [@Pearl_et_al_2016, pp. 29]. Specifically, it expresses the joint distribution as a product of conditional probability distributions (CPDs) of the type $P(child \cond parents)$. This property is formally known as the *Bayesian Network factorization* (BNF) [@Pearl_et_al_2016, pp. 29; @Neal_2020, pp. 21]. Equation [-@eq-mat28] presents the joint distribution of the sample data and partitions it into smaller CPDs derived from the SCM structure. -->

<!-- $$ -->
<!-- \begin{aligned} -->
<!--   P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\ -->
<!--   &= P(O_{R} \;|\; S, C, D_{R}) \cdot P(D_{R} \;|\; T_{IA}, B_{JK}) \cdot P(T_{IA} \;|\; T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \;|\; X_{I}, e_{I}) \\ -->
<!--   & \quad \cdot P(B_{JK} \;|\; B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \;|\; Z_{J}, e_{J}) \cdot P(S) \cdot P(C) \cdot P(X_{IA}) \cdot P(X_{I}) \\  -->
<!--   & \quad \cdot P(Z_{JK}) \cdot P(Z_{J}) \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J}) -->
<!-- \end{aligned} -->
<!-- $$ {#eq-mat28} -->

<!-- 1. How to derive an stat model from an SCM? aboutn the outcome distribution -->
Researchers can easily derive a statistical model to address violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a) using only [SCM -@fig-cj15_scm]. This derivation is possible because the structural equations $F$ encode probabilistic information [@Pearl_et_al_2016], which researchers can replace with suitable probabilistic assumptions. As a starting point, given the dichotomous nature of CJ outcomes, [model -@fig-cj16_stat] assumes that $O_{R}$ follows a Bernoulli distribution[^6].

Next, following the conventions of Generalized Linear Latent And Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the model links the observed outcome to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$. Notice that the sampling and comparison mechanisms, $S$ and $C$, act as element-wise multipliers of $D_{R}$ (denoted by the symbol $\odot$), directly incorporating key assessment design features relevant to CJ experiments into the statistical formulation. Researchers can omit these vectors when only the observed CJ sample is available.

<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution. ... For any model whose graph is acyclic, the joint distribution of the variables in the model is given by the product of the conditional distributions P(child|parents) over all the "families" in the graph. [@Pearl_et_al_2016, pp. 29] -->

[^6]: The binomial distribution--including its special case, the Bernoulli distribution-- represent a maximum entropy distribution for binary events [@McElreath_2020, pp. 34]. This means that when only two un-ordered outcomes are possible, and the expected numbers of each type of event are assumed to be constant, then the binomial distribution is the one that is most consistent with these constraints [@McElreath_2020, pp. 310]. For a detailed discussion of the binomial as a maximum entropy distribution, see @McElreath_2020 [chap. 10.1.2].

@@

::: {#fig-cj16 layout-ncol=2}

::: {#fig-cj16_stat fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
$$

::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & := f_{O}(D_{R}, S, C) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
  e_{IA} & \dsep e_{JK} 
\end{aligned}
$$

[SCM -@fig-cj15_scm]
:::

Comparative judgment model, SCM and statistical model assuming different discriminal dispersions for the student's traits
:::

<!-- 2. Continuing with the probabilistic assumptions -->
Then, [model -@fig-cj16_stat] defines $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, which represent their underlying written-quality trait. This difference also includes the corresponding repeated judge biases, $B_{JK}[j, k]$. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, then we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] ] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat51}
This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases). For clarity, the square brackets in $D_{R}$ only indicate the relevant indices within each trait vector; they do not represent any data subsetting.

@@

Next, the model defines $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length), and the texts' idiosyncratic errors $e_{IA}$. The model assumes that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$, defined as a proportion of $1$. This standardization establishes the scale of the latent trait, as it is required by latent variable models [@Depaoli_2021].

@@

Additionally, the model defines $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$. As before, each element of $e_{I}$ is $iid$ following a Normal distribution with a mean of zero. However, the standard deviation, $s_{XI}$, varies depending on the teaching method to which each student belongs. Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$  This constraint effectively addresses concerns about the equal dispersion assumption raised in @sec-theory-issue1a.

@@

Similarly, the model defines $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., the number of judgments a judge makes), and judgment-specific idiosyncratic errors $e_{JK}$. The model assumes that each element of $e_{JK}$ is $iid$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].

>>>>

<!-- commands for d-separation -->
\newcommand{\dsep}{\:\bot\:}
\newcommand{\ndsep}{\:\not\bot\:}

@@

<!-- 1. The SCMs -->
Before incorporating the first theoretical principles of Thurstone's theory, it is essential to further define SCMs. SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Pearl_et_al_2016; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other endogenous and exogenous variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

<!-- Formally, a structural causal model consists of two sets of variables U and V, and a set of functions f that assigns each variable in V a value based on the values of the other variables in the model. ... The variables in U are called exogenous variables, meaning, roughly, that they are external to the model; we choose, for whatever reason, not to explain how they are caused. The variables in V are endogenous. Every endogenous variable in a model is a descendant of at least one exogenous variable. Exogenous variables cannot be descendants of any other variables, and in particular, cannot be a descendant of an endogenous variable; they have no ancestors and are represented as root nodes in graphs. If we know the value of every exogenous variable, then using the functions in f , we can determine with perfect certainty the value of very endogenous variable. [@Pearl_et_al_2016, pp. 26-27] -->

@@

<!-- Every SCM is associated with a graphical causal model, referred to informally as a “graphical model” or simply “graph.” Graphical models consist of a set of nodes representing the variables in U and V, and a set of edges between the nodes representing the functions in f. [@Pearl_et_al_2016, pp. 27] -->

<!-- Every SCM has an associated graph G, usually called its causal diagram. In the types of models we consider here, which do not exhibit cycles, the causal diagram will be a directed acyclic graph (DAG). [@Cinelli_et_al_2020, pp. 15] -->

@@

<!-- Do not depict the variances, nor the sample sizes. Or if you do explain that it is with illustrative purposes and then remove them to make the reading of the DAGs easier. -->

@@

<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
C -> {Y}
e_{A} -> {C, Y} 
this means that e_{A} is a confounder?,  
then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?. Does this mean that random algorithms per se will always be unbiased?. -->

@@

<!-- recording: Sven and Tine 25.04.10; time: 00:16:30 - 00:19:00 -->

<!-- 1. start by defining the comparison vector  -->
As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model (see equation ([-@eq-mat27])). The vector contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in equation ([-@eq-mat26]).
$$
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat27}

<!-- 2. What is the effect of the comparison mechanism -->
The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$. It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism. Moreover, it depicts $C$ as a causal factor influencing $O_{R}$. As with the sampling mechanism, a square encloses $C$ to indicate that it is a conditioned variable--that is, one that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*. In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli. Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp. 267].

@@

Notably, [DAG -@fig-cj14_dag] also shows that $C$ is independent of all other variables in the model. This independence implies that the conceptual model represented by the DAG applies exclusively to Random Allocation Comparative Designs [@Bramley_2015] or Incomplete Block Designs [@Lawson_2015]. In these designs, every repeated judgment has an equal probability of being included in the sample.

@@

<!-- 3. preparing to transform the conceptual model into an statistical one -->
Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15. This reformulation produces a model that applies directly to a sample of comparative data. In this version, the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$ are omitted, and $O_{R}$ inherits the structural equation $f_{O}$ that originally defined $O^{cp}_{R}$. Moreover, the definition of $O_{R}$ now reflects its direct dependence on the discriminal difference $D_{R}$ and the sample and comparison mechanisms, $S$ and $C$.

<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population. For example, if I have the distribution of the population outcome, and it is a Bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->

<!-- 4. segway for next section -->
In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model. These extensions account for judge biases (see @sec-theory-issue1b), reflect the hierarchical structure of stimuli (see @sec-theory-issue1b), and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2). However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a. Because this issue involves statistical assumptions about the distribution of the discriminal process, it requires a statistical model, which we develop in the next section.

# From SCMs to a statistical model  {#sec-statistical}

<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
<!-- recording: Sven and Tine 25.03.28; time: 00:33:55 - 00:37:50 -->
<!-- recording: Sven and Tine 25.03.31; time: 00:38:50 - 00:47:00 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:00:00 - 00:01:10 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:01:10 - 00:01:35 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:01:35 - 00:16:30 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:19:00 - 00:26:40 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:52:00 - 00:53:40 -->

<!-- 1. How to derive an stat model from an SCM? -->
With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a). This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with appropriate functions and probabilistic assumptions [@Pearl_et_al_2016]. Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]). For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent. Here, $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
$$
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
$$ {#eq-mat51}

<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution. For any model whose graph is acyclic, the joint distribution of the variables in the model is given by the product of the conditional distributions P(child|parents) over all the "families" in the graph. [@Pearl_et_al_2016, pp. 29] -->

<!-- The functions F = {fz, fx, fy} are called structural equations, and each function represents a causal process that assigns to its respective endogenous variable a value based on the values of the other variables. [@Cinelli_et_al_2020, pp. 15] -->

[^6]: This re-expression is possible because the *chain rule* of probability and the *Bayesian Network Factorization (BNF)* property. For further details, see @Pearl_et_al_2016 and @Neal_2020.

<!-- We will denote the parents of a node X with pa(X). We’ll use an even simpler shorthand when the nodes are ordered so that we can denote the i^{th} node by X_{i}; in that case, we will also denote the parents of X_{i} by pa_{i} [@Neal_2020, pp. 19] Definition 3.1 (Bayesian Network Factorization) Given a probability distribution P and a DAG G, P factorizes according to G if: P(x_{1}, \dots, x_{n}) = \prod_{i}P(x_{i} | pa_{i}). [@Neal_2020, pp. 21] -->

<!-- For any model whose graph is acyclic, the joint distribution of the variables in the model is given by the product of the conditional distributions P(child | parents) over all the "families" in the graph. Formally, we write this rule as P(x_{1}, \dots, x_{n}) = \prod_{i}P(x_{i} | pa_{i}) where pa_{i} stands for the values of the parents of variable X_{i}, and the product i runs over all i, from 1 to n. [@Pearl_et_al_2016, pp. 29] -->

<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline below. First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$. Note that the probability distributions for the sample and comparison mechanisms, $P(S)$ and $P(C)$, included in the probabilistic model [-@fig-cj16_prob], are omitted from the statistical model [-@fig-cj16_stat]. We exclude these probabilities from the statistical definition because, although they are part of the joint distribution of the data, they do not influence the dependence between $O_{R}$ and $D_{R}$, since $S$ and $C$ are governed by random selection designs[^8].

[^7]: The binomial distribution--including its special case, the Bernoulli distribution-- represent a maximum entropy distribution for binary events [@McElreath_2020, pp. 34]. This means that when only two un-ordered outcomes are possible, and the expected numbers of each type of event are assumed to be constant, then the binomial distribution is the one that is most consistent with these constraints [@McElreath_2020, pp. 310]. For a detailed discussion of the binomial as a maximum entropy distribution, see @McElreath_2020 [chap. 10.1.2].

@@

[^8]: Randomization ensures that data—and, by extension, an estimator—satisfies several key identification properties, such as common support, no interference, and consistency. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $S$), resulting in spurious associations between the two [@Everitt_et_al_2010]. Randomization addresses this issue by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself. For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.

<!-- 2. Continuing with the assumptions -->
Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$. The discriminal processes represent the underlying written-quality traits of the texts. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat52}
This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases). For clarity, the square brackets in $D_{R}$ only indicate the relevant indices within each trait vector; they do not represent any data subsetting.

@@

<!-- Judges discrimination ability can be integrated in the models if we assume that the bias is not additive (as in Bramley explanation) but multiplicative. In the latter case the bias is not eliminated by taking just the difference. True, but it is false to think that because the biases are additive they get just eliminated. For instance, if we assume that the bias is additive but different for each stimuli, this means that the DAG should be: {T_{IA}, B_{JK}} -> O_{R}, but mathematically you can write it as it is above  -->

::: {#fig-cj16 layout-ncol=3}

::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & := f_{O}(D_{R}, S, C) \\ 
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
  e_{IA} & \dsep e_{JK} \\ \\
\end{aligned}
$$

::: {#fig-cj16_prob fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\
  & P( e_{I} ) \\
  & P( e_{J} ) \\
  & P( e_{IA} ) \\  
  & P( e_{JK} )
\end{aligned}
$$

Probabilistic model
:::

::: {#fig-cj16_stat fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
$$

Comparative judgment model, SCM, probabilistic and statistical model assuming different discriminal dispersions for the student's traits
:::

Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$. We assume that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$. The parameter $p_{IA}$ is defined as a proportion of 1 to establish the scale of the latent trait, which is required in any latent variable model [@Depaoli_2021]. Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed. This result indicates that our model reverts to the original assumptions for the discriminal processes of stimuli proposed by @Thurstone_1927b (see @tbl-thurstone_cases).

@@

Fourth, we define $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$. As before, we assume that each element of $e_{I}$ is $iid$ Normal with a mean of zero. However, we consider the standard deviation, $s_{XI}$, varies depending on the teaching method group to which each student belongs. Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$. Note that this parametrization of $e_{I}$ relaxes the equal dispersion assumption for the stimuli, effectively addressing the concerns raised in @sec-theory-issue1a.

@@

Fifth, we define $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments a judge makes affect the assessment of quality), and judgment-specific idiosyncratic errors $e_{JK}$. We assume that each element of $e_{JK}$ is $iid$ Normal with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].

Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$. Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs. For instance, if $Z_{J} = 1, \dots, 3$ represents three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
